Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-09-24 hazelnut-popl17.tex
working on #6

\def\OPTIONConf{1}%
%\newif\iftr\trfalse
\newif\iftr\trtrue
%\documentclass[12pt]{article}
\RequirePackage{etex}
\documentclass[preprint,9pt]{sigplanconf}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage{multicol}
\usepackage{extarrows}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{light-gray}{gray}{0.9}
\usepackage{soul}
\setulcolor{red}
\usepackage{mathpazo}
\usepackage{colortab}
\usepackage{url}
\usepackage{todonotes}
\usepackage{listings}
\lstset{tabsize=2,
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont}
\usepackage{microtype}
\sloppy
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

@@

%\usepackage{balance} %% Balance last page's two columns

@@

\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink

@@

\usepackage{todonotes}
\usepackage{xcolor}
%\usepackage{adjustbox}

@@

%\titlebanner{}        % These are ignored unless
\preprintfooter{Draft}   % 'preprint' option specified.

\title{Hazelnut: A Bidirectionally Typed \\ Structure Editor
 Calculus}
% \title{Hazelnut:\\A Bidirectionally Typed Structure Editor Calculus}

%\authorinfo{\vspace{-2px}}{}{}
%\authorinfo{~}{~}{\vspace{-10px}}
%\authorinfo{Cyrus Omar \and Jonathan Aldrich}
%         {Carnegie Mellon University}
%         {\{comar, aldrich\}@cs.cmu.edu}
\iftr
\authorinfo{
        Cyrus Omar$^1$
        \and Ian Voysey$^1$
        \and Michael Hilton$^2$
        \and Jonathan Aldrich$^1$
        \and Matthew A. Hammer$^3$
}
{
 $^1$~Carnegie Mellon University
 \and
 $^2$~Oregon State University
 \and
 $^3$~University of Colorado Boulder\vspace{-10px}
}
{
%\email{comar@cs.cmu.edu}\\
%\texttt{http://www.cs.cmu.edu/\homedir comar/}
%\email{hiltonm@eecs.oregonstate.edu}\\
}
\else
\authorinfo{~}{~}{\vspace{-10px}}
\fi

%
%\authorrunning{Omar et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
%\email{Matthew.Hammer@colorado.edu}

\maketitle
\begin{abstract}
% Programmers typically construct and manipulate well-typed expressions  only indirectly, as text that must pass through a parser and type\-checker. Not all text survives this journey. In particular, text that arises transiently, or when the programmer makes a mistake, is often malformed or ill-typed.
% Not all text that arises during the programming process survives the journey through a parser and typechecker. In particular, text that arises transiently during the editing process, or when the programmer has made a mistake, is often malformed or ill-typed.
% Contending with malformed program text or well-formed but meaningless syntax trees is difficult for programmers and their tools alike.

%  \emph{Structure editors} have long promised to alleviate these burdens by exposing only edit actions that  cause sensible changes to the program structure.
% Existing designs for structure editors, however, are complex and somewhat \emph{ad hoc}. They also focus primarily on syntactic well-formedness, so programs can still be left semantically meaningless as they are being constructed.
% Structure editors (e.g. Scratch) eliminate the possibility of syntax errors.

@@

%Formally, Hazelnut is a bidirectionally typed lambda calculus extended with \emph{holes}, a \emph{focus model} (based on Huet's zipper) and an \emph{action model}.
\end{abstract}

@@

% \clearpage

% \section{Dynamics of Hazelnut}\label{sec:dynamics}

% \subsection{Complete Expressions}
% ... canonical forms and type safety with e complete in it ...

% \subsection{Holes as TODOs}
% ... add an e err and define new type safety ...

% \subsection{Indeterminate Expressions}
% \input{dynamics}

% \clearpage

\section{Extending Hazelnut}\label{sec:extending}
%% \todo{write this section} Add sum types.
%% H-expression
%% Z-expression + Erasure rules
%% Statics
%% Movement actions
%% Construct actions
\input{sumtypes}

@@

\section{Related Work and Discussion}\label{sec:rw}
%\subsection{Structure Editors}
\subsection{Structure Editors}
Syntactic structure editors have a long history -- the Cornell Program Synthesizer~\cite{teitelbaum_cornell_1981} was first introduced in 1981. Novice programmers have been a common target for structure editors. For example,
GNOME~\cite{garlan_gnome:_1984} was developed to teach programming to undergraduates.
Alice~\cite{Conway:2000:ALL:332040.332481} is a 3-D programming language with an integrated structure editor for teaching novice CS undergraduate students. Scratch~\cite{Resnick:2009:SP:1592761.1592779} is a structure editor targeted at children ages 8 to 16.  TouchDevelop \cite{tillmann_touchdevelop:_2011} incorporates a structure editor for programming on touch-based devices, and is used to teach high school students. An implementation of Hazelnut might be useful in teaching students about the typed lambda calculus, though that has not been our explicit aim with this work.

@@

Some structure editor generators do operate on formal or semi-formal definitions of an underlying language. For example, the Synthesizer Generator~\cite{Reps:1984:SG:390010.808247} allows the user to define an attribute grammar-based language implementation that then can be used to generate a structured editor. CENTAUR~\cite{Borras:1988:CS:64140.65005} produces a language specific environment from a user defined formal specification of a language. Barista is a programmatic toolkit for building structure editors \cite{ko_barista:_2006}. mbeddr is built on top of the commercial JetBrains MPS framework for constructing structure editors \cite{voelter2011language,DBLP:journals/software/VoelterWK15}. These systems do not give a semantics to the edit states of the structure editor itself, or maintain non-trivial edit invariants, as Hazelnut does.
%These early systems were developed  of the systems are rooted in the type-theoretic tradition.

@@

The Graphite system used an informal model of typed holes in Java to allow library providers to associate interactive code generation interfaces with types \cite{Omar:2012:ACC:2337223.2337324}. %RThis system was limited in that these interfaces could not themselves manipulate terms.

More generally, the topic of type-directed program synthesis an active area of research, e.g. \cite{DBLP:conf/pldi/OseraZ15}. By maintaining static well-definedness throughout the editing process, Hazelnut provides researchers interested in editor-integrated type-directed program synthesis  with a formal foundation upon which to build.% such systems could be integrated directly into editors.

@@

One interesting avenue of future work is to elaborate expression holes to CMTT's closures, i.e. CMTT terms of the form $\mathsf{clo}(u; \text{id}(\Gamma))$ where $u$ is a unique metavariable associated with each hole and $\text{id}(\Gamma)$ is the explicit identity substitution. This would allow us to evaluate expressions with holes such that the closure ``accumulates'' substitutions explicitly. When evaluation gets ``stuck'' (as it can, for CMTT does not define a dynamics equipped with a notion of progress under a non-empty modal context), it would then be possible for the programmer to choose holes from amongst the visible holes (which may have been duplicated) to edit in their original context. Once finished, the CMTT hole instantiation operation, together with a metatheorem that establishes that reduction commutes with instantiation, would enable an ``edit and resume'' feature with a clear formal basis. This notion of reduction commuting with instantiation has also been studied in other calculi \cite{DBLP:journals/entcs/Sands97}. Being able to edit a running program also has connections to less formal work on ``live programming'' interfaces \cite{burckhardt2013s,lamdu}. %, e.g.

@@

% By keeping the program in both a structurally and semantically well-defined state at all times, Hazelnut allows users to avoid premature commitment~\cite{green1996usability}.
% By inserting a hole, the user can leave certain parts of the program unfinished, \todo{What is the best word to define this concept?  In progress, unfinished, incomple?} and yet still in a well defined type state.
% This also enables progressive evaluation~\cite{green1996usability}, because unfinished solutions are also are well-defined at all times, thus enabling tools to provide evaluation of unfinished solutions.\todo{integrate this}

@@

% We already discussed a connection to gradual typing \cite{Siek06a}. We hope to explore this connection more thoroughly. In particular, it may be possible to better support exploratory and live programming by allowing even programs with holes in them to execute as long as those holes are only in the type portions, by deferring to the semantics given in work on gradual typing.

% It may also be possible to give a dynamics to incomplete expressions. Prior work on staged evaluation suggests that there may be a connection to modal logic, viewing holes as quantifying over all possible terms that may fill them \cite{DBLP:journals/jacm/DaviesP01}. In developing a dynamic semantics, we will also need to handle terms like $\hhole{\hehole}$ and
% $\hhole{\hhole{\hexp}}$. In our semantics given here, we eliminated them as they came up in a somewhat \emph{ad hoc} manner. We have not yet
% explored an equational theory for terms with holes, but intend to once our
% formalization effort is more mature.\todo{fix bib}

@@

\clearpage
% We recommend abbrvnat bibliography style.
\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.
\bibliography{bibliography}
%\softraggedright
%P. Q. Smith, and X. Y. Jones. ...reference text...

>>>>

\def\OPTIONConf{1}%
%\newif\iftr\trfalse
\newif\iftr\trtrue
%\documentclass[12pt]{article}
\RequirePackage{etex}
\documentclass[preprint,9pt]{sigplanconf}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{wasysym}
\usepackage{multicol}
\usepackage{extarrows}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{light-gray}{gray}{0.9}
\usepackage{soul}
\setulcolor{red}
\usepackage{mathpazo}
\usepackage{colortab}
\usepackage{url}
\usepackage{todonotes}
\usepackage{listings}
\lstset{tabsize=2,
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont}
\usepackage{microtype}
\sloppy
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

@@

@@

% known bug: http://tex.stackexchange.com/questions/1522/pdfendlink-ended-up-in-different-nesting-level-than-pdfstartlink
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}

@@

\usepackage{todonotes}
\usepackage{xcolor}

@@

\preprintfooter{Draft}   % 'preprint' option specified.

\title{Hazelnut: A Bidirectionally Typed \\ Structure Editor
 Calculus}
\iftr
\authorinfo{
        Cyrus Omar$^1$
        \and Ian Voysey$^1$
        \and Michael Hilton$^2$
        \and Jonathan Aldrich$^1$
        \and Matthew A. Hammer$^3$
}
{
 $^1$~Carnegie Mellon University
 \and
 $^2$~Oregon State University
 \and
 $^3$~University of Colorado Boulder\vspace{-10px}
}
{
%\email{comar@cs.cmu.edu}\\
%\texttt{http://www.cs.cmu.edu/\homedir comar/}
%\email{hiltonm@eecs.oregonstate.edu}\\
}
\else
\authorinfo{~}{~}{\vspace{-10px}}
\fi

\maketitle
\begin{abstract}

@@

\end{abstract}

@@

\section{Extending Hazelnut}\label{sec:extending}
\input{sumtypes}

@@

\section{Related Work and Discussion}\label{sec:rw}
\subsection{Structure Editors}
Syntactic structure editors have a long history -- the Cornell Program
Synthesizer~\cite{teitelbaum_cornell_1981} was first introduced in
1981. Novice programmers have been a common target for structure
editors. For example, GNOME~\cite{garlan_gnome:_1984} was developed to
teach programming to undergraduates.
Alice~\cite{Conway:2000:ALL:332040.332481} is a 3-D programming language
with an integrated structure editor for teaching novice CS undergraduate
students. Scratch~\cite{Resnick:2009:SP:1592761.1592779} is a structure
editor targeted at children ages 8 to 16.
TouchDevelop \cite{tillmann_touchdevelop:_2011} incorporates a structure
editor for programming on touch-based devices, and is used to teach high
school students. An implementation of Hazelnut might be useful in teaching
students about the typed lambda calculus, though that has not been our
explicit aim with this work.

@@

Some structure editor generators do operate on formal or semi-formal
definitions of an underlying language. For example, the Synthesizer
Generator~\cite{Reps:1984:SG:390010.808247} allows the user to define an
attribute grammar-based language implementation that then can be used to
generate a structured editor. CENTAUR~\cite{Borras:1988:CS:64140.65005}
produces a language specific environment from a user defined formal
specification of a language. Barista is a programmatic toolkit for building
structure editors \cite{ko_barista:_2006}. mbeddr is built on top of the
commercial JetBrains MPS framework for constructing structure
editors \cite{voelter2011language,DBLP:journals/software/VoelterWK15}. These
systems do not give a semantics to the edit states of the structure editor
itself, or maintain non-trivial edit invariants, as Hazelnut does.

@@

The Graphite system used an informal model of typed holes in Java to allow
library providers to associate interactive code generation interfaces with
types \cite{Omar:2012:ACC:2337223.2337324}.

More generally, the topic of type-directed program synthesis an active area
of research, e.g. \cite{DBLP:conf/pldi/OseraZ15}. By maintaining static
well-definedness throughout the editing process, Hazelnut provides
researchers interested in editor-integrated type-directed program synthesis
with a formal foundation upon which to build.

@@

One interesting avenue of future work is to elaborate expression holes to
CMTT's closures, i.e. CMTT terms of the form
$\mathsf{clo}(u; \text{id}(\Gamma))$ where $u$ is a unique metavariable
associated with each hole and $\text{id}(\Gamma)$ is the explicit identity
substitution. This would allow us to evaluate expressions with holes such
that the closure ``accumulates'' substitutions explicitly. When evaluation
gets ``stuck'' (as it can, for CMTT does not define a dynamics equipped
with a notion of progress under a non-empty modal context), it would then
be possible for the programmer to choose holes from amongst the visible
holes (which may have been duplicated) to edit in their original
context. Once finished, the CMTT hole instantiation operation, together
with a metatheorem that establishes that reduction commutes with
instantiation, would enable an ``edit and resume'' feature with a clear
formal basis. This notion of reduction commuting with instantiation has
also been studied in other
calculi \cite{DBLP:journals/entcs/Sands97}. Being able to edit a running
program also has connections to less formal work on ``live programming''
interfaces \cite{burckhardt2013s,lamdu}.

@@

% By keeping the program in both a structurally and semantically
% well-defined state at all times, Hazelnut allows users to avoid premature
% commitment~\cite{green1996usability}.  By inserting a hole, the user can
% leave certain parts of the program unfinished, \todo{What is the best
% word to define this concept?  In progress, unfinished, incomple?} and yet
% still in a well defined type state.  This also enables progressive
% evaluation~\cite{green1996usability}, because unfinished solutions are
% also are well-defined at all times, thus enabling tools to provide
% evaluation of unfinished solutions.\todo{integrate this}

@@

% We already discussed a connection to gradual typing \cite{Siek06a}. We
% hope to explore this connection more thoroughly. In particular, it may be
% possible to better support exploratory and live programming by allowing
% even programs with holes in them to execute as long as those holes are
% only in the type portions, by deferring to the semantics given in work on
% gradual typing.

% It may also be possible to give a dynamics to incomplete
% expressions. Prior work on staged evaluation suggests that there may be a
% connection to modal logic, viewing holes as quantifying over all possible
% terms that may fill them \cite{DBLP:journals/jacm/DaviesP01}. In
% developing a dynamic semantics, we will also need to handle terms like
% $\hhole{\hehole}$ and $\hhole{\hhole{\hexp}}$. In our semantics given
% here, we eliminated them as they came up in a somewhat \emph{ad hoc}
% manner. We have not yet explored an equational theory for terms with
% holes, but intend to once our formalization effort is more
% mature.\todo{fix bib}

@@

\clearpage

\bibliographystyle{abbrvnat}
% The bibliography should be embedded for final submission.
\bibliography{bibliography}
